Red Wine Data Analysis by Marco Rizk
========================================================



```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(GGally)
library(scales)
library(memisc)
library(grid)
library(gridExtra)
```


```{r echo=FALSE, Load_the_Data}
# Load the Data
red_wine <- read.csv('wineQualityReds.csv')
```

> The "red wine quality" dataset explores the chemical factors affecting wine \
quality as represented by different taste experts.

# Univariate Plots Section


```{r echo=FALSE, Univariate_Plots}
#investigating the data structure and a general summary of the variables
str(red_wine)
summary(red_wine)
head(red_wine)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#investigating univariateplots
p1 <- qplot(data = red_wine, x=quality) +geom_bar() +   
  scale_x_continuous(breaks = seq(3,8,1))

p2 <- qplot(data = red_wine, x=alcohol)
p3 <- qplot(data = red_wine, x=sulphates)+
  scale_x_continuous(limits=c(min(red_wine$sulphates), 
                              quantile(red_wine$sulphates, 0.95)))
p4 <- qplot(data = red_wine, x=pH)
p5 <- qplot(data = red_wine, x=density)
p6 <- qplot(data = red_wine, x=total.sulfur.dioxide)
p7 <- qplot(data = red_wine, x=free.sulfur.dioxide)
p8 <- qplot(data = red_wine, x=chlorides) +
  scale_x_continuous(limits=c(min(red_wine$chlorides), 
                              quantile(red_wine$chlorides, 0.96)))
p9 <- qplot(data = red_wine, x=residual.sugar)+
  scale_x_continuous(limits=c(min(red_wine$residual.sugar), 
                              quantile(red_wine$residual.sugar, 0.9)))
p10 <- qplot(data = red_wine, x=citric.acid)
p11 <- qplot(data = red_wine, x=volatile.acidity )
p12 <- qplot(data = red_wine, x=fixed.acidity )

grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,p12,ncol=3)

```


#comments on plots

1. The quality plot shows that it follows normal distribution with the mean and\
median around 5 to 6
2. The chlorides and residual sugar plots show a narrow band for most wine samples\
so they are less significant in further analysis
3. the rest of the variables need to be investigated further for their effect on \
the quality
4. Alcohol is between 8 and 14% as expected of wine
5. sulphates have a left skewed distrubtion with the bulk between 0.5 and 1
6. The bulk of the pH is between 3 and 3.5 which is slightly acidic as expected \
of wines


# Univariate Analysis

### What is the structure of your dataset?
The dataset consists of 1599 obsevations, structured into a 13 variable table\
which are:
Sample no. , fixed acidity, volatile acitdity, citric acid, residual sugar, \
chlorides, free sulfur oxide , total sulfur oxide, density, pH, sulphates, \
alcohol, quality.

### What is/are the main feature(s) of interest in your dataset?
The main feature is "quality". we need to investigate what factors affect \
the quality


### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
putting the chlorides and residual sugar aside. I think acidity has rhe major\
effect on qulaity (whether fixed or volatile) which would refect on pH too\
Also we need to investigate how different acids affect quality.

### Did you create any new variables from existing variables in the dataset?
No the exisitng variables are sufficent for the investigation.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
No operations were done so far. some of the variables had outliers that will be\
handled when needed.


# Bivariate Plots Section



```{r echo=FALSE, message=FALSE, warning=FALSE}
ggcorr(red_wine[,-1
                ], 
       nbreaks = 4, 
       label = TRUE, 
       label_size = 4, 
       label_color = "white", 
       size = 4)
```

#Observations:

1. The factor affecting quality are: alcohol, volatile acidity(negatively) \
and sulphates
2. Alcohol is negatively correlated with denisty which is normal to expect
3. Sulphates are correlated with chlorides
4. pH is negatively correlated with fixed acidity whis is expected as lower \
pH indicates higher acidity
5. density is correlated with fixed acidity
6. citric acid is correlated with acidity

```{r echo=FALSE, message=FALSE, warning=FALSE}
#factos affecting quality
p1 <- ggplot(aes(x=alcohol,y=quality), data=red_wine) +
  geom_jitter(alpha=0.2) + 
  geom_boxplot(alpha=0.2,color="orange")+
  geom_smooth(method = "lm")
p2 <- ggplot(aes(x=volatile.acidity,y=quality), data=red_wine) +
  geom_jitter(alpha=0.2) +
  geom_boxplot(alpha=0.2,color="orange")+
  geom_smooth(method = "lm")
p3 <- ggplot(aes(x=sulphates,y=quality), data=red_wine) +
  geom_jitter(alpha=0.2) +
  geom_boxplot(alpha=0.2,color="orange")+
  geom_smooth(method = "lm")
grid.arrange(p1,p2,p3,ncol=3)
```

# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

The factors that affect the quality the most are Alcohol and Volatile acidity\
and sulphates

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
1. Fixed acidity and pH are negatively  correlated which is expected as lower\
pH indicates higher acidity
2. Also Fixed acidity is strongly correlated to density
3. Alcohol and density are negatively correlated which is expected

### What was the strongest relationship you found?
The strongest was between density and fixed acidity 

# Multivariate Plots Section



```{r echo=FALSE, Multivariate_Plots}
ggplot(aes(x=alcohol,y=sulphates,color=factor(quality)), data=red_wine)+
  geom_point() + scale_colour_brewer(palette = "RdYlGn")


```

It can be observed that best quality wines are those of high alcohol content\
and sulphates between 0.5 and 1

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x=alcohol,y=volatile.acidity,color=factor(quality)), data=red_wine)+
  geom_point() + scale_colour_brewer(palette = "RdYlGn")

```

Another observation would be that higher quality wine are the ones associated\
with higher alcohol content and voaltile acidity less than nearly 0.6

```{r echo=FALSE, message=FALSE, warning=FALSE}
m1 <- lm(formula = I(quality) ~ I(alcohol), data = red_wine)
m2 <- update(m1, ~ . + volatile.acidity)
m3 <- update(m2, ~ . + log10(sulphates))
m4 <- update(m3, ~ . + log10(chlorides))
m5 <- update(m4, ~ . + pH)
m6 <- update(m5, ~ . + citric.acid)

mtable(m1,m2,m3,m4,m5,m6, sdigits = 3)
```


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \

investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
1. an increase in alcohol percent and sulphates between 0.5 and 1 are associates\
with better wine quality
2.higher quality wine are the ones associated with higher alcohol conenet and\
voaltile acidity less than nearly 0.6


### Were there any interesting or surprising interactions between features?

Although wine quality was correlated with high % of alcohol/
some wine showed good quality while being low on alcohol due/ 
to the effect of higher sulphates

------

# Final Plots and Summary



### Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(quality), data = red_wine) +
  geom_bar() +
  labs(title = 'Histogram of Wine Quality',
       x = 'Quality of Wine',
       y = 'Number of Samples') +
  scale_x_continuous(breaks = seq(3,8,1))

```

### Description One

The majority of wine samples have a medium quality of 5 to 6 which makes it \
difficult to build a reliable model since the samples for both high and low\
quality wines are low in count.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(factor(quality), alcohol), data = red_wine) +
  geom_jitter( alpha = .3)  +
  
  geom_smooth(aes(quality-2, red_wine$alcohol), 
              method = "lm", size=1, color = 'orange') +
  labs(x = "Alcohol %",
       y = "Quality",
       title = "Quality by Alcohol Content")

```

### Description Two

Alcohol has the strongest correlation to wine quality as observed by the\
linearally fitted model

### Plot Three

```{r}

ggplot(aes(sulphates, alcohol, color = factor(quality)), 
       data = red_wine) +
  geom_jitter(size = 2, alpha = 1/2) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  labs(x = "Sulphates",
       y = "Alcohol %",
       title = "Quality by Sulphates and Alcohol")


```

### Description Three

The best quality wines have high values for both Alcohol percentage and \
Sulphate concentration implying that High alcohol contents and high sulphate\
concentrations together seem to produce better wines

#Reflection

The following results can be concluded from the analysis:

1. Wine quality is mainly affected by Alcohol percent also some other factor\
such as, sulphates, chlorides affect the quality slightly
2. some other factors seem to be correlated together like desnity/fixed acidity\
or citric acid and pH which can be attributed to some physical phenomna
3. Generated model for predicting the quality of wines is weak and only able to\
predict 37% of the sample with multiple parameters for prediction

##Recommendation

The data needs to be expanded to include more sample for the low quality and\
high quality wines (which are rare in provided data) for better understanding\
of the factors affecting the quality and also for better quality prediction.

